<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.3.7/css/bootstrap.min.css">
<style>
  .right{
    float: right;
  }
  img{
    display:block;
    margin: 0 auto;
  
  }
  pre{
    overflow-x: hidden;
    overflow-y: hidden;
  }
  body{
    margin-top: 20px;
    margin-bottom: 20px;
  }
  .col-left{
    float: left;
    width: 50%;
  }
  .col-right{
    float: right;
    width: 50%;
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(ggplot2)
require(RColorBrewer)
require(MASS)
require(plyr)
require(scales)
```

<div class='jumbotron'>
  <h2 class='display-3 text-uppercase'>Are Fewer Children Born on Weekends?</h2>
  <h4 class='right text-uppercase'>Analysis by Brian Weinfeld</h4>
</div>

<div class='page-header text-uppercase'>
  <h3>The Claim</h3>
</div>
<div class='well'>
  <p><span class='lead'>D</span>uring a recent discussion with some proud new parents, they told me an anecdote told to them by their doctor. The doctor had told them that more children were born during the week than during the weekend. This idea immediately intrigued me as logically the day of the week should have no affect on when a child is born. I decided to investigate.
  </p>
</div>

```{r}
birthData <- read.table(file='https://raw.githubusercontent.com/brian-cuny/rassignment3/master/Births.csv', header=TRUE, sep=',', stringsAsFactors=FALSE)
head(birthData,n=10)
```

<div class='well'>
  <p><span class='lead'>I</span> began by pulling data on births in the United States from 1969 to 1988. The raw data showed that the number of births on any given day in this set was appoximately normal centered around a mean of 9649 with a standard deviation of 1127.315 births.</p>
</div>

<div class='col-left'>
```{r}
summary(birthData$births)
```
</div>
<div class='col-right'>
```{r}
sd(birthData$births)
```
</div>
<div class='clearfix'></div>

```{r}
ggplot(birthData, aes(x=births)) + geom_histogram(aes(y=..density.., fill=..count..), bins=50) + scale_fill_gradient('Count', low='#DCDCDC', high='#7C7C7C') + stat_function(fun=dnorm, color='red', args=list(mean=mean(birthData$births), sd=sd(birthData$births)), aes(size=2), show.legend=FALSE) + theme(legend.position='none') + scale_y_continuous('Proportion', labels=scales::comma) + scale_x_continuous('Births', labels=scales::comma) + labs(title='Frequency of Births in Single Day (1969-1988)', caption='Data From: http://vincentarelbundock.github.io/Rdatasets/') 
```

<div class='well'>
  <p><span class='lead'>I</span> removed the unneeded columns and then agregated the data to show the total number of births for each day of the week for each year. In this data set 1 through 5 represented Monday through Friday while 6 and 7 represented Saturday and Sunday respsectively.</p>
</div>

```{r}
workingData <- birthData[,c(5,9,3)]
combinedBirthData <- aggregate(births ~ day_of_week + year, workingData, sum)[,c(2,1,3)]
head(combinedBirthData,n=10)
```

<div class='well'>
  <p><span class='lead'>I</span> then created a scatter plot of the data comparing births with day of the week and drew a line through the mean of each day's data. Visual inspection of the graph seemed to indicate that there was a significant dip in births on Saturday and Sunday. I also plotted each year seperately in case one year's data was skewing the overall results. Visual inspection of that data supported the initial observation.</p>
</div>

```{r}
original <- ggplot(combinedBirthData, aes(x=factor(day_of_week), y=births)) + geom_point(aes(color=factor(year))) + 
  labs(title='Births By Day of Week (1969-1988) With Trend Line', caption='Data From: http://vincentarelbundock.github.io/Rdatasets/',x='Day Of Week', y='Births', color='Year')

original + stat_summary(aes(y=births,group=1,size=1.5), fun.y=mean, geom='line',group=1, show.legend=FALSE) + scale_x_discrete(labels=c('Mon', 'Tues','Wed','Thurs','Fri','Sat','Sun')) 
```

```{r}
original + facet_wrap(~year) + theme(legend.position='none') + scale_x_discrete(labels=c('M','T','W',"Th",'F','Sa','Su'))
```

<div class='page-header text-uppercase'>
  <h3>Statistical Analysis</h3>
</div>
<div class='well'>
  <p><span class='lead'>A</span>lthough visual inspection seemed to indicate a difference in birth rates it is important to perform a test to ensure that this is supportred statistically. An ANOVA Test would identify if there were a statistically significant difference between the number of births for each day of the week across all the data. However, I first wanted to ensure that the data had no outliers. I created a boxplot showing the range of births for each day of the week. This confirmed there were no outliers that needed to be addressed prior to the ANOVA test.</p>
</div>

```{r}
ggplot(combinedBirthData, aes(x=factor(day_of_week), y=births)) + geom_boxplot() + scale_x_discrete(labels=c('Mon', 'Tues','Wed','Thurs','Fri','Sat','Sun')) + 
  labs(title='Births By Day of Week (1969-1988)', caption='Data From: http://vincentarelbundock.github.io/Rdatasets/',x='Day Of Week', y='Births') 
```

```{r}
summary(aov(births ~ day_of_week, data=workingData))
```

<div class='page-header text-uppercase'>
  <h3>Initial Conclusion</h3>
</div>
<div class='alert alert-success'>
  <p>With a p-value of appoximately $2\times10^{-16}$ I can confidently say that there was a statistically significant difference between the birth rates of babies when compared to the day of the week.</p>
</div>
<div class='well'>
  <p><span class='lead'>F</span>urther research yielded a perfectly reasonable explanation for my results. Improvements in medical technology have allowed doctors to schedule deliveries in advance and induce labor at the appropriate time. Combine this fact with the fact that doctors are more likely to take the weekends off and this leads to a logical explanation for the data. Doctors take weekends off so they schedule deliveries for the weekdays.</p>
</div>
<div class='page-header text-uppercase'>
  <h3>Growing Trend?</h3>
</div>
<div class='well'>
  <p><span class='lead'>W</span>ith that mystery solved I turned to another observation I noticed from visual inspection of the data. It appeared that the discrepency between weekday and weekend births are growing more pronounced as the years went by. Had this trend been growing during the examined time frame?</p>
  <p>I began by writing a function to aid in creating a new data frame that identied the proportion of births that occurred during the week days.</p>
</div>
  
```{r}
weekDayFunction <- function(data){
  sum(data[which(data[,2] %in% 1:5),]$births) / sum(data$births)
}

ratioData <- ddply(workingData, .variables='year', .fun=weekDayFunction)
names(ratioData) <- c('year', 'proportion_weekday')
```

<div class='col-left'>
```{r}
head(ratioData, n=10)
```
</div>
<div class='col-right'>
```{r}
tail(ratioData, n=10)
```
</div>
<div class='clearfix'></div>

<div class='well'>
  <p><span class='lead'>I</span> then plotted each year's proportion of week day births with a regression line. The regresssion line appears to show a clear growing trend in this proportion. I performed a least-square regression analysis.</p>
</div>

```{r}
ggplot(ratioData, aes(x=year, y=proportion_weekday)) + geom_point(aes(color=proportion_weekday)) + scale_color_gradient(low='blue', high='red') + labs(title='Proportion of Births during Weekday (1969-1988)', x='Year', y='Proprotion of Babies Born', caption='Data From: http://vincentarelbundock.github.io/Rdatasets/') + theme(legend.position='none') + scale_y_continuous(labels=scales::percent) + geom_smooth(method='lm')
```

```{r}
ratioData$year <- ratioData$year - 1969
summary(lm(formula = proportion_weekday ~ year, data=ratioData))
cor(ratioData$proportion_weekday, ratioData$year)
```

<div class='page-header text-uppercase'>
  <h3>Final Conclusion</h3>
</div>
$$\widehat{proportion}=.0981(year)+73.88$$
<div class='well'>
  <p><span class='lead'>T</span>he least squares regression line above where $year$ represents the number of the years since 1969 and $\widehat{proportion}$  represents the proportion of weekday births indicates that there is an increase of appoximately $\frac{1}{10}$th of a percentage point per year (or 1 percentage point per decade) in the rate of weekday births compared to weekend births. With a p-value of appoximately $1.56\times10^{-9}$ it can safely be concluded that this is a statistically significant trend. The correlation $r=0.9349669$ supports this conclusion.</p>
</div>
<div class='alert alert-success'>
  It can be stated with strong confidence that weekday births have been steadily increasing in the United States in the years 1969 to 1988. It can also be stated with strong confidence that there is a statitically significant increase in the number of babies born during the week than during the weekend.
</div>
<div class='well'>
  <p>There are a number of follow up questions I would be interested in pursuing:</p>
  * Why did the rate of doctors scheduling weekday births increase?
  * Is the rate still increasing or did it level off?
  * What is the predominate cause for weekend births?
</div>










